---
title: "Portfolio 2"
output: github_document
---

## Part5: Attention Data Analysis

Now that the attention data has been successfully extracted and saved as a CSV file, the next step is to perform statistical analyses and visualize the results. As described in the introduction, each participant completed multiple trials/blocks, so observations are correlated within participants. Participants themselves can be seen as a random sample. Therefore, participants are treated as a random effect to account for within-participant dependence and to allow individual differences in baseline levels. 

```{r load-analysis-packages, warning = FALSE, message=FALSE}
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyverse) 
library(tidyr)
library(R.matlab)
```

### 5.1 Data Preparation

First, let's load the attention data and recode variables into meaningful labels.

```{r attention-data-prep, warning = FALSE, message=FALSE}
att <- read.csv("attention_data.csv")

# recoding
att$most_condition <- factor(ifelse(att$block_type %in% c(1, 3), "mostly_shift", "mostly_hold"))
att$trial_type <- factor(ifelse(att$trial_type == 2, "shift", "hold"))
att$Congruence <- factor(ifelse(att$congruence == 1, "congruent", "incongruent"))
att$phase <- factor(ifelse(att$run %in% c(1, 3), "encoding", "retrieval"))
att$subject_id <- factor(att$subid)

str(att)
```

### 5.2 Encoding Phase Analysis

For the encoding phase, we fit a 2 (most_condition: mostly_shift vs. mostly_hold) × 2 (trial_type: shift vs. hold) linear mixed-effects model with participant as random intercept.

```{r encoding-model, warning = FALSE, message=FALSE}
encoding_data <- att %>% filter(phase == "encoding")

# Fit mixed-effects model
encoding_model <- lmer(RT ~ most_condition * trial_type + (1 | subject_id), 
                       data = encoding_data)

summary(encoding_model)$varcor
anova(encoding_model, type = 3)
```

A more intuition way to show is :
```{r encoding-means, warning = FALSE, message=FALSE}
encoding_means <- encoding_data %>%
  group_by(most_condition, trial_type) %>%
  summarize(
    mean_RT = round(mean(RT, na.rm = TRUE), 3),
    se_RT = round(sd(RT, na.rm = TRUE) / sqrt(n()), 4),
    n = n(),
    .groups = "drop"
  )
encoding_means
```

The key finding here is the **most_condition × trial_type interaction**: Under the mostly_hold condition, there is a large shift cost (shift trials are much slower than hold trials). However, under the mostly_shift condition, this difference nearly disappears, as participants are better adapted to shifting.

```{r encoding-plot, warning = FALSE, message=FALSE, fig.width=7, fig.height=5}
ggplot(encoding_means, aes(x = trial_type, y = mean_RT, 
                            color = most_condition, shape = most_condition, group = most_condition)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = mean_RT - se_RT, ymax = mean_RT + se_RT), width = 0.1) +
  geom_text(aes(label = mean_RT), vjust = -1.2, size = 3.5) +
  scale_color_manual(values = c("mostly_hold" = "#2196F3", "mostly_shift" = "#9C27B0"),
                     labels = c("Mostly Hold", "Mostly Shift")) +
  scale_shape_manual(values = c("mostly_hold" = 16, "mostly_shift" = 17),
                     labels = c("Mostly Hold", "Mostly Shift")) +
  labs(x = "Trial Type", y = "Response Time (s)", 
       color = NULL, shape = NULL) +
  scale_x_discrete(labels = c("hold" = "Hold", "shift" = "Shift")) +
  theme_minimal() +
  theme(legend.position = c(0.15, 0.95),
        legend.background = element_rect(fill = "white", color = "black"),
        text = element_text(size = 12))
```


### 5.3 Retrieval Phase Analysis

For the retrieval phase, congruence (whether encoding and retrieval share the same block type) becomes relevant. So we fit a 2 (most_condition) × 2 (trial_type) × 2 (Congruence) mixed-effects model.

```{r retrieval-model, warning = FALSE, message=FALSE}
retrieval_data <- att %>% filter(phase == "retrieval")

# Fit three-way mixed-effects model
retrieval_model <- lmer(RT ~ most_condition * trial_type * Congruence + (1 | subject_id), 
                        data = retrieval_data)

# Random effects
summary(retrieval_model)$varcor

# ANOVA table
anova(retrieval_model, type = 3)
```

Let's examine the cell means:
```{r retrieval-means, warning = FALSE, message=FALSE}
retrieval_means <- retrieval_data %>%
  group_by(most_condition, trial_type, Congruence) %>%
  summarize(
    mean_RT = round(mean(RT, na.rm = TRUE), 3),
    se_RT = round(sd(RT, na.rm = TRUE) / sqrt(n()), 4),
    n = n(),
    .groups = "drop"
  )
retrieval_means
```

The retrieval phase yields a significant three-way interaction (most_condition × trial_type × Congruence). This is an interesting result. Congruence appears to slightly modulate the magnitude of the effect under certain conditions. For incongruent trials, participants actually seem to adapt better to shifting: transitioning from holding to shifting (or vice versa) appears to make shifting trials themselves more fluent. This may reflect a form of abstract rule-based attentional flexibility.

```{r retrieval-plot, warning = FALSE, message=FALSE, fig.width=10, fig.height=5}
ggplot(retrieval_means, aes(x = trial_type, y = mean_RT, 
                             color = most_condition, shape = most_condition, group = most_condition)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = mean_RT - se_RT, ymax = mean_RT + se_RT), width = 0.1) +
  geom_text(aes(label = mean_RT), vjust = -1.2, size = 3.5) +
  facet_wrap(~ Congruence) +
  scale_color_manual(values = c("mostly_hold" = "#2196F3", "mostly_shift" = "#9C27B0"),
                     labels = c("Mostly Hold", "Mostly Shift")) +
  scale_shape_manual(values = c("mostly_hold" = 16, "mostly_shift" = 17),
                     labels = c("Mostly Hold", "Mostly Shift")) +
  labs(x = "Trial Type", y = "Response Time (s)", 
       color = NULL, shape = NULL) +
  scale_x_discrete(labels = c("hold" = "Hold", "shift" = "Shift")) +
  theme_minimal() +
  theme(legend.position = c(0.08, 0.95),
        legend.background = element_rect(fill = "white", color = "black"),
        text = element_text(size = 12),
        strip.text = element_text(size = 13, face = "bold"))
```


## Part 6: Memory Data Extraction

As mentioned at the beginning, the memory component involves more complex extraction steps. In this part, I will translate the memory extraction pipeline from my Matlab code into R. The key idea is: memory is tested at the **pair level** — run 1 (encoding) pairs with run 2 (retrieval), and run 3 pairs with run 4. During the retrieval run, participants are tested on items they saw during encoding. 

The critical variables to extract from memory data are:
  
## Memory Data Extraction

However, this memory dataset had several issues during data collection.

+ the task scripts were not fully consistent across participants. As a result, for the earlier and later portions of the sample, the response keys were stored in different cells.

+ some participants forgot the instructions and chose to skip trials by pressing the space bar.

+ in some trials, the location of the correct answer was not stored properly.

+ the correct answer information was also saved in different places depending on the script version. For some participants, it was stored in `testorder`, whereas for others it was stored in `testselection`.

+ memory extraction differs fundamentally from attention extraction. In the attention task, each block can be analyzed independently. In contrast, for the memory task, encoding and retrieval do not occur in the same block. Therefore, we need to extract two runs simultaneously, analyze the retrieval run, and then determine whether each memory trial is congruent or incongruent relative to its encoding run.

Therefore, the logic of the memory extraction function is relatively straightforward. We first read the data from Run 2 and Run 4. Then, we check whether the second list in `testselection` contains `0`. If it does, we extract `testorder` from the corresponding 12th or 14th list. Meanwhile, for `testselection`, we only read the raw key codes as they were stored. For Run 2, the responses are read from the second list of `testselection`; for Run 4, they are read from the fourth list. After that, we record each trial index and which block it belongs to, and then compute whether each trial is congruent or incongruent.

### 6.1 Memory Extraction Function

<span style="color: deepskyblue;"> This code is extremely complex. I used GPT to help debug several scenarios, but it still didn't work. I had no choice but to add numerous `if` statements. But in fact, I feel that the time I spend doing this is far more than that of manually organizing the data. *(Good news is People make mistakes so it worth, Bad news is Of course people and GPT-written code do too lol)*</span>

```{r memory-extract-function, warning = FALSE, message=FALSE}
library(R.matlab)
library(dplyr)
library(tibble)

get_p_field <- function(mat_obj, field_name) {
  tryCatch({
    mat_obj$p[field_name, 1, 1][[1]]
  }, error = function(e) NULL)
}

flatten_nested <- function(x) {
  if (is.null(x)) return(NULL)
  if (is.atomic(x)) return(x)
  if (is.list(x)) return(unlist(lapply(x, flatten_nested), recursive = TRUE, use.names = FALSE))
  return(NULL)
}


extract_candidate <- function(x, idx) {
  tryCatch({
    if (is.null(x) || !is.list(x)) return(NULL)
    if (idx > length(x)) return(NULL)
    out <- as.numeric(flatten_nested(x[[idx]]))
    out <- out[!is.na(out)]
    if (length(out) == 0) return(NULL)
    out
  }, error = function(e) NULL)
}

is_valid_order <- function(x) {
  if (is.null(x) || length(x) == 0) return(FALSE)
  x <- x[!is.na(x)]
  if (length(x) == 0) return(FALSE)
  if (any(x == 0)) return(FALSE)
  if (!all(x %in% c(1, 2))) return(FALSE)
  TRUE
}

# 根据 run 和数据结构尝试多个候选位置
choose_testorder <- function(ts, to, ret_run) {
  sel_idx <- ifelse(ret_run == 2, 2,
                    ifelse(ret_run == 4, 4, NA))
  
  if (is.na(sel_idx)) return(NULL)
  
  ts2 <- extract_candidate(ts, 2)
  has_zero_in_second <- !is.null(ts2) && any(ts2 == 0, na.rm = TRUE)
  

  preferred_order_idx <- ifelse(ret_run == 2, 12,
                                ifelse(ret_run == 4, 14, NA))
  
  candidates <- list()
  

  if (has_zero_in_second && !is.na(preferred_order_idx)) {
    candidates[[paste0("to_", preferred_order_idx)]] <- extract_candidate(to, preferred_order_idx)
  }
  
  candidates[[paste0("to_", sel_idx)]] <- extract_candidate(to, sel_idx)
  
  if (!is.na(preferred_order_idx)) {
    candidates[[paste0("ts_", preferred_order_idx)]] <- extract_candidate(ts, preferred_order_idx)
  }

  candidates[[paste0("ts_", sel_idx)]] <- extract_candidate(ts, sel_idx)
  

  for (nm in names(candidates)) {
    cand <- candidates[[nm]]
    if (is_valid_order(cand)) {
      message("Using TestOrder source: ", nm)
      return(cand)
    }
  }
  
  for (nm in names(candidates)) {
    cand <- candidates[[nm]]
    if (!is.null(cand) && length(cand) > 0) {
      cand_clean <- cand[cand %in% c(1, 2)]
      if (length(cand_clean) > 0) {
        message("Fallback TestOrder source: ", nm, " (filtered to 1/2 only)")
        return(cand_clean)
      }
    }
  }
  
  NULL
}

extract_memory_run <- function(p_enc, p_ret, pair_num) {
  
  subject_id  <- suppressWarnings(as.numeric(get_p_field(p_enc, "PID")))
  enc_run     <- suppressWarnings(as.numeric(get_p_field(p_enc, "run")))
  ret_run     <- suppressWarnings(as.numeric(get_p_field(p_ret, "run")))
  enc_version <- suppressWarnings(as.numeric(get_p_field(p_enc, "Version")))
  ret_version <- suppressWarnings(as.numeric(get_p_field(p_ret, "Version")))
  
  enc_block <- ifelse(enc_version %in% c(1, 3), "mostly_shift", "mostly_hold")
  ret_block <- ifelse(ret_version %in% c(1, 3), "mostly_shift", "mostly_hold")
  congruence <- ifelse(enc_version == ret_version, "congruent", "incongruent")
  
  ts <- get_p_field(p_ret, "testselection")
  to <- get_p_field(p_ret, "testorder")
  
  if (is.null(ts) || is.null(to)) return(NULL)
  if (!is.list(ts) || !is.list(to)) return(NULL)
  
  sel_idx <- ifelse(ret_run == 2, 2,
                    ifelse(ret_run == 4, 4, NA))
  if (is.na(sel_idx)) return(NULL)
  

  raw_selection <- extract_candidate(ts, sel_idx)
  if (is.null(raw_selection) || length(raw_selection) == 0) return(NULL)

  raw_testorder <- choose_testorder(ts, to, ret_run)
  if (is.null(raw_testorder) || length(raw_testorder) == 0) return(NULL)
  

  raw_testorder <- raw_testorder[raw_testorder %in% c(1, 2)]
  if (length(raw_testorder) == 0) return(NULL)
  
  n <- min(length(raw_selection), length(raw_testorder))
  raw_selection <- raw_selection[1:n]
  raw_testorder <- raw_testorder[1:n]
  
  tibble(
    SubjectID = subject_id,
    Pair = pair_num,
    EncodingRun = enc_run,
    RetrievalRun = ret_run,
    EncodingBlock = enc_block,
    RetrievalBlock = ret_block,
    Congruence = congruence,
    Trial = seq_len(n),
    TestOrder = raw_testorder,
    Selection = raw_selection
  )
}

extract_memory_subject <- function(run1, run2, run3, run4) {
  pair1 <- extract_memory_run(run1, run2, pair_num = 1)
  pair2 <- extract_memory_run(run3, run4, pair_num = 2)
  bind_rows(pair1, pair2)
}
```

### 6.2 Batch Processing for Memory Data

Similar to the attention batch processing, we loop through all participants' files and extract memory data.

```{r memory-batch, warning = FALSE, message=FALSE}
all_files <- list.files("data", pattern = "Within_run.*\\.mat$", full.names = TRUE)
all_files <- sort(all_files)
all_memory <- list()

for (i in seq(1, length(all_files), by = 4)) {
  
  if ((i + 3) > length(all_files)) next
  
  run1 <- tryCatch(readMat(all_files[i]),   error = function(e) NULL)
  run2 <- tryCatch(readMat(all_files[i+1]), error = function(e) NULL)
  run3 <- tryCatch(readMat(all_files[i+2]), error = function(e) NULL)
  run4 <- tryCatch(readMat(all_files[i+3]), error = function(e) NULL)
  
  if (is.null(run1) || is.null(run2) || is.null(run3) || is.null(run4)) next
  
  memory_result <- extract_memory_subject(run1, run2, run3, run4)
  
  if (!is.null(memory_result) && nrow(memory_result) > 0) {
    subid <- suppressWarnings(as.numeric(get_p_field(run1, "PID")))
    all_memory[[as.character(subid)]] <- memory_result
  }
}

memory_data <- bind_rows(all_memory)

write.csv(memory_data, "memory_data.csv", row.names = FALSE)
head(memory_data)
```

## Part7: Memory Data Analysis

### 7.1 Accuracy Summary

First, let's compute per-participant accuracy rates, aggregated by RetrievalBlock and Congruence.

```{r memory-accuracy-summary, warning=FALSE, message=FALSE}
memory_data <- memory_data %>%
  mutate(
    Correct = case_when(
      Selection == 66 ~ NA_real_,
      Selection == 53 & TestOrder == 1 ~ 1,
      Selection == 59 & TestOrder == 2 ~ 1,
      TRUE ~ 0
    )
  )

table(memory_data$Correct, useNA = "ifany")
memory_data <- memory_data %>%
  mutate(
    SubjectID = factor(SubjectID),
    RetrievalBlock = factor(RetrievalBlock),
    Congruence = factor(Congruence)
  )

# Per-participant accuracy by condition
memory_summary <- memory_data %>%
  group_by(SubjectID) %>%
  filter(!any(is.na(Correct))) %>%
  group_by(SubjectID, RetrievalBlock, Congruence) %>%
  summarise(
    accuracy = mean(Correct), 
    n_trials = n(),
    .groups = "drop"
  )
```



### 7.2 Mixed-Effects ANOVA on Accuracy 

I first fit a linear mixed-effects model on the aggregated accuracy rates.

```{r memory-lmer, warning=FALSE, message=FALSE}
library(lme4)
library(lmerTest)

memory_model_agg <- lmer(
  accuracy ~ RetrievalBlock * Congruence + (1 | SubjectID),
  data = memory_summary
)

summary(memory_model_agg)
anova(memory_model_agg, type = 3)
```

```{r memory-descriptive, warning=FALSE, message=FALSE}
library(dplyr)

# Descriptive statistics by condition
memory_desc <- memory_summary %>%
  group_by(RetrievalBlock, Congruence) %>%
  summarise(
    mean_accuracy = mean(accuracy, na.rm = TRUE),
    sd_accuracy = sd(accuracy, na.rm = TRUE),
    n_subjects = n(),
    .groups = "drop"
  )

memory_desc
```

### 7.3 Memory Accuracy Visualization

```{r memory-plot, warning=FALSE, message=FALSE, fig.width=7, fig.height=5}
library(emmeans)
library(ggplot2)

# Estimated marginal means
memory_emm <- emmeans(memory_model_agg, ~ RetrievalBlock * Congruence)
memory_emm_df <- as.data.frame(memory_emm)

ggplot(memory_emm_df,
       aes(x = Congruence,
           y = emmean,
           color = RetrievalBlock,
           shape = RetrievalBlock,
           group = RetrievalBlock)) +
  geom_point(size = 3) +
  geom_line(linewidth = 1) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.1) +
  geom_text(aes(label = round(emmean, 3)), vjust = -1.5, size = 3.5) +
  scale_color_manual(
    values = c("mostly_hold" = "#2196F3", "mostly_shift" = "#9C27B0"),
    labels = c("Mostly Hold", "Mostly Shift")
  ) +
  scale_shape_manual(
    values = c("mostly_hold" = 16, "mostly_shift" = 17),
    labels = c("Mostly Hold", "Mostly Shift")
  ) +
  scale_y_continuous(
    labels = scales::percent_format(),
    limits = c(0.7, 1.0)
  ) +
  labs(
    x = "Congruence",
    y = "Memory Accuracy (Estimated Marginal Means)",
    title = "Memory Accuracy: Mixed-Effects ANOVA",
    color = "Retrieval Block",
    shape = "Retrieval Block"
  ) +
  theme_minimal() +
  theme(
    text = element_text(size = 12),
    legend.position = "top"
  )
```


