---
title: " Visualizing Brain Regions from Literature"
author: "Holland Sun"
output: github_document
---

I know there are some websites that can visualize which brain region is related to certain cognitive function. Like [neurosynth](https://neurosynth.org) by searching for "attention," we can see the distribution of brain regions associated with it, and even some effect sizes (as this website is actually design for meta-analyses). But here, I'd like to do something a bit more interesting (Of course more simple) specifically foucus on attention flexibility. It's not rigorous science, but it's a fun exercise in text mining and visualization.

## Portfolio Goals

+ Learn how to import literature data (exported from PubMed) into R
+ Learn text mining and extract brain region names from abstracts
+ Use the `ggseg` package to visualization


## Intro

Neuroimaging studies often report which brain regions are activated during specific cognitive tasks. If you read enough papers on "attentional flexibility," you'll notice certain regions popping up again and again: prefrontal cortex, anterior cingulate, parietal areas, etc. But what if we wanted to quantify that impression? 

A proper approach would be a coordinate-based meta-analysis (like ALE or MKDA), where you extract MNI/Talairach coordinates from each study and compute statistical convergence maps. But as I mention before such tools are serious for research. So what I am doing here is much lazier : we're just counting how many times each brain region name appears across abstracts. 


## Part 1: Geting data

### Search literature 

Before we write any R code, we need data. If we just search "attentional flexibility," we'll get a mix of behavioral studies, clinical studies, and neuroimaging studies. We want to focus on human neuroimaging, so a reasonable PubMed query would be something like:

**("attentional flexibility" OR "attentional shifting" OR "attentional set shifting" OR "attention switching" OR "task switching") AND (fMRI OR neuroimaging OR "brain activation" OR PET)**

In practice, I went to PubMed and exported the results. PubMed's CSV export gives you columns like PMID, Title, Authors, Abstract, etc. 

For reproducibility, I saved both files in the `data/` folder:

+ `pubmed_metadata.csv` — the CSV export (for clean metadata)
+ `pubmed_abstracts.txt` — the PubMed format export (for abstracts)

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(stringr)
```

## Part 2: Text mining 

The CSV file has clean metadata:

```{r load-csv, warning=FALSE, message=FALSE}
metadata <- read_csv("data/pubmed_metadata.csv")
glimpse(metadata)
cat("Columns:", paste(names(metadata), collapse = ", "))
```
For the basic information They looks all good (although we won't use them in our following analysis). 

Now For the most tricky part. The PubMed text file looks something like this:

```
1. Neuroscience. 2017 Mar 14;345:243-255...

Cognitive inflexibility in Obsessive-Compulsive Disorder.

Gruner P(1), Pittenger C(2).

Author information:
...

Obsessive-Compulsive Disorder (OCD) is characterized by maladaptive patterns of
repetitive, inflexible cognition and behavior...

DOI: 10.1016/j.neuroscience.2016.07.030
PMCID: PMC5288350
PMID: 27491478 [Indexed for MEDLINE]
```

Each entry ends with a `PMID:` line, and entries are separated by blank lines. The abstract is the longest continuous paragraph in each entry and it comes after the author info and before the copyright/DOI lines. So My strategy here is spliting the file into individual entries, then extract the PMID and the abstract from each.

```{r parse-abstracts}
raw_text <- readLines("data/pubmed_abstracts.txt", warn = FALSE)
full_text <- paste(raw_text, collapse = "\n")

# Split into individual entries — each starts with a number like "1. " or "2. "
entries <- str_split(full_text, "(?=\n\\d+\\. )")[[1]]
entries <- entries[nchar(trimws(entries)) > 0]

cat("Number of entries found:", length(entries), "\n")
```

Now I need to extract the PMID and abstract from each entry. This requires some regex (happy to know this term in English, in Chinese it has a very weird name). The abstract is typically the longest block of text between the author information section and the copyright/DOI section.

```{r extract}
extract_pubmed_entry <- function(entry) {
  # Extract PMID 
  pmid_match <- str_extract(entry, "PMID:\\s*(\\d+)")
  pmid <- str_extract(pmid_match, "\\d+")
  
  lines <- str_split(entry, "\n")[[1]]
  lines <- trimws(lines)
  
  # The structure of each PubMed entry is:
  #   1. Citation info (journal, year, doi...)
  #   [blank line]
  #   2. Title
  #   [blank line]
  #   3. Author names
  #   [blank line]
  #   4. Author information
  #   [blank line]        
  #   5. Abstract text
  #   [blank line]
  #   6. Copyright / DOI / PMID
  
  abstract_lines <- c()
  in_abstract <- FALSE
  in_author_info <- FALSE
  has_author_info <- any(str_detect(lines, "^Author information:"))
  
  # Find where author names are (line with parenthetical affiliations like "Smith J(1)")
  author_line_idx <- 0
  
  for (i in seq_along(lines)) {
    line <- lines[i]
    
    # -- Phase 1: detect "Author information:" section
    if (str_detect(line, "^Author information:")) {
      in_author_info <- TRUE
      next
    }
    
    # -- Phase 2: skip all affiliation lines until blank line
    if (in_author_info && !in_abstract) {
      if (line == "") {
        # blank line after affiliations = abstract starts next
        in_author_info <- FALSE
        in_abstract <- TRUE
        next
      }
      # still in affiliations, skip
      next
    }
    
    # -- For entries WITHOUT Author information:
    # The abstract follows the author names after a blank line.
    # Author names typically contain parenthetical numbers like "(1)"
    if (!has_author_info && !in_abstract) {
      if (str_detect(line, "\\(\\d+\\)") && i > 2) {
        author_line_idx <- i
      }
      # After the author name block, look for blank line then abstract
      if (author_line_idx > 0 && line == "" && i > author_line_idx) {
        in_abstract <- TRUE
        next
      }
      next
    }
    
    # -- Phase 3: collect abstract lines
    if (in_abstract) {
      # Stop at DOI, copyright, PMID, or empty-then-metadata
      if (str_detect(line, "^(DOI:|PMCID:|PMID:|©|Copyright|Published by|Conflict of interest)")) break
      if (line == "") {
        # A blank line inside abstract could be end of abstract.
        # Check if next non-blank line is metadata
        remaining <- lines[(i+1):min(i+3, length(lines))]
        remaining <- remaining[remaining != ""]
        if (length(remaining) > 0 && 
            str_detect(remaining[1], "^(DOI:|PMCID:|PMID:|©|Copyright|Published by|Conflict of interest)")) {
          break
        }
        # otherwise it might be a paragraph break within abstract, keep going
      }
      if (line != "") {
        abstract_lines <- c(abstract_lines, line)
      }
    }
  }
  
  abstract <- paste(abstract_lines, collapse = " ")
  abstract <- str_squish(abstract)
  
  tibble(PMID = as.integer(pmid), Abstract = abstract)
}
```

To be honest, this part is not perfect. PubMed's text format is... inconsistent. When I first use the simple strategy I only get around 50 abstract. Becasue Some entries have author information blocks, some don't. Some have copyright notices, some don't. It is a bit messy. So I turn to ChatGpt to help me find the reason, Although he still know nothing, I check some possibility and let him help me on those `if` condition code. It's too messy

<span style="color: deepskyblue;">*I then find there is a R package called `rentrez`, which can directly  pull abstracts from the NCBI API using PMIDs. But I am satisfied with just using text here)*</span>

```{r run function, warning=FALSE}
abstracts_list <- map(entries, possibly(extract_pubmed_entry, 
                                         tibble(PMID = NA_integer_, Abstract = "")))
abstracts_df <- bind_rows(abstracts_list)
abstracts_df <- abstracts_df %>% filter(!is.na(PMID))
papers <- metadata %>%
  inner_join(abstracts_df, by = "PMID")

papers %>%
  select(PMID, Abstract) %>%
  slice(1:5)
```

We now have a tidy dataframe with metadata AND abstracts. 

## Part 3: Counting

So since we decide just count the frequency of those regions, we need a **dictionary* to search for in the abstracts. This is where it gets a bit tricky, because brain regions have many aliases. "dlPFC" and "dorsolateral prefrontal cortex" refer to the same area. And some region names are substrings of others ("frontal cortex" is inside "prefrontal cortex"). So we need to be somewhat careful with our dictionary.

I'll create a dictionary that maps various text patterns to standardized region names. And I'll also map them to the `ggseg` atlas region labels

```{r brain-dictionary}


brain_dict <- tribble(
  ~pattern,                            ~region_standard,                    ~ggseg_label,
  "dorsolateral prefrontal",           "Dorsolateral PFC",                  "rostralmiddlefrontal",
  "dlPFC",                             "Dorsolateral PFC",                  "rostralmiddlefrontal",
  "DLPFC",                             "Dorsolateral PFC",                  "rostralmiddlefrontal",
  "middle frontal gyrus",              "Middle Frontal Gyrus",              "caudalmiddlefrontal",
  "inferior frontal gyrus",            "Inferior Frontal Gyrus",            "parsopercularis",
  "inferior frontal cortex",           "Inferior Frontal Gyrus",            "parsopercularis",
  "right inferior frontal",            "Inferior Frontal Gyrus",            "parsopercularis",
  "inferior frontal junction",         "Inferior Frontal Junction",         "parsopercularis",
  "ventrolateral prefrontal",          "Ventrolateral PFC",                 "parstriangularis",
  "vlPFC",                             "Ventrolateral PFC",                 "parstriangularis",
  "VLPFC",                             "Ventrolateral PFC",                 "parstriangularis",
  "superior frontal gyrus",            "Superior Frontal Gyrus",            "superiorfrontal",
  "superior frontal cortex",           "Superior Frontal Gyrus",            "superiorfrontal",
  "medial prefrontal",                 "Medial PFC",                        "medialorbitofrontal",
  "medial frontal cortex",             "Medial Frontal",                    "medialorbitofrontal",
  "orbitofrontal cortex",              "Orbitofrontal Cortex",              "lateralorbitofrontal",
  "orbitofrontal",                     "Orbitofrontal Cortex",              "lateralorbitofrontal",
  "anterior cingulate",                "Anterior Cingulate",                "rostralanteriorcingulate",
  "ACC",                               "Anterior Cingulate",                "rostralanteriorcingulate",
  "posterior cingulate",               "Posterior Cingulate",               "posteriorcingulate",
  "cingulate cortex",                  "Cingulate Cortex",                  "caudalanteriorcingulate",
  "cingulate gyrus",                   "Cingulate Cortex",                  "caudalanteriorcingulate",
  "superior parietal",                 "Superior Parietal Lobule",          "superiorparietal",
  "intraparietal sulcus",              "Intraparietal Sulcus",              "superiorparietal",
  "inferior parietal",                 "Inferior Parietal Lobule",          "inferiorparietal",
  "angular gyrus",                     "Angular Gyrus",                     "inferiorparietal",
  "supramarginal gyrus",               "Supramarginal Gyrus",              "supramarginal",
  "temporoparietal junction",          "Temporoparietal Junction",          "supramarginal",
  "precuneus",                         "Precuneus",                         "precuneus",
  "insula",                            "Insula",                            "insula",
  "anterior insula",                   "Insula",                            "insula",
  "frontal eye field",                 "Frontal Eye Fields",                "precentral",
  "precentral gyrus",                  "Precentral Gyrus",                  "precentral",
  "precentral sulcus",                 "Precentral Gyrus",                  "precentral",
  "premotor cortex",                   "Premotor Cortex",                   "precentral",
  "supplementary motor",               "SMA / Pre-SMA",                     "paracentral",
  "pre-supplementary motor",           "SMA / Pre-SMA",                     "paracentral",
  "pre-SMA",                           "SMA / Pre-SMA",                     "paracentral",
  "fusiform gyrus",                    "Fusiform Gyrus",                    "fusiform",
  "fusiform",                          "Fusiform Gyrus",                    "fusiform",
  "lateral occipital",                 "Lateral Occipital",                 "lateraloccipital",
  "visual cortex",                     "Visual Cortex",                     "lateraloccipital",
  "occipital cortex",                  "Occipital Cortex",                  "lateraloccipital",
  "posterior parietal cortex",         "Posterior Parietal",                "superiorparietal",
  "posterior parietal",                "Posterior Parietal",                "superiorparietal",
  "temporal cortex",                   "Temporal Cortex",                   "middletemporal",
  "middle temporal gyrus",             "Middle Temporal Gyrus",             "middletemporal",
  "superior temporal gyrus",           "Superior Temporal Gyrus",           "superiortemporal",
  "superior temporal sulcus",          "Superior Temporal Gyrus",           "superiortemporal",
  "inferior temporal",                 "Inferior Temporal Gyrus",           "inferiortemporal",
  "postcentral gyrus",                 "Postcentral Gyrus",                "postcentral",
  "somatosensory",                     "Postcentral Gyrus",                "postcentral",
  "cuneus",                            "Cuneus",                            "cuneus",
  "lingual gyrus",                     "Lingual Gyrus",                     "lingual",
  "entorhinal",                        "Entorhinal Cortex",                "entorhinal",
  "parahippocampal",                   "Parahippocampal Gyrus",            "parahippocampal",
  "hippocampus",                       "Hippocampus",                       "parahippocampal"
)

```

the mapping to `ggseg` labels is approximate. The Desikan-Killiany atlas parcellates the cortex into about 34 regions per hemisphere. And it doesn't have "dlPFC" as a label. So I'm mapping "dlPFC" to "rostralmiddlefrontal," which is the closest DK parcel. There should be lot of other issues, but it is just a practice, and Thanks agian for GPT helps me do such dirty work.

```{r count-regions}
papers <- papers %>%
  mutate(full_text = paste(Title, Abstract, sep = " "))

# For each dictionary pattern
region_counts <- brain_dict %>%
  rowwise() %>%
  mutate(
    count = sum(str_detect(papers$full_text, 
                           regex(pattern, ignore_case = TRUE)))
  ) %>%
  ungroup()

# Aggregate 
region_summary <- region_counts %>%
  group_by(ggseg_label) %>%
  summarise(
    display_name = region_standard[which.max(count)],
    total_mentions = sum(count),
    .groups = "drop"
  ) %>%
  arrange(desc(total_mentions))

region_summary
```

## Part 4: Mapping onto ggseg Brain Atlas

The `ggseg` package lets you plot brain atlas parcellations using ggplot2 syntax. We need to join our frequency data with the atlas region names.

```{r brain-plot, warning=FALSE, message=FALSE, fig.width=10, fig.height=12}
library(ggseg)
library(sf)

plot_data <- region_summary %>%
  rename(region = ggseg_label) %>%
  select(region, total_mentions)

# ggseg version handling: dk might be a function (2.0+) or data (1.6.x)
dk_atlas <- if (is.function(dk)) dk() else dk

atlas_sf <- as.data.frame(dk_atlas) %>% st_as_sf()

merged <- atlas_sf %>%
  left_join(plot_data, by = "region")

ggplot(merged) +
  geom_sf(aes(fill = total_mentions), color = "gray30", linewidth = 0.2) +
  scale_fill_gradient(low = "lightyellow", high = "red",
                      na.value = "gray90",
                      name = "Mention\nFrequency") +
  labs(title = "Brain Regions Mentioned in Attentional Flexibility Literature",
       subtitle = "Based on word frequency in abstracts (NOT a real meta-analysis!)",
       caption = paste0("Data: PubMed search (n=", nrow(papers), 
                        " papers, searched Mar 8, 2026) | Atlas: Desikan-Killiany")) +
  theme_void() +
  theme(
    plot.title = element_text(hjust = 0.5, size = 14, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, size = 10, color = "gray40"),
    plot.caption = element_text(hjust = 0.5, size = 8, color = "gray50"),
    legend.position = "right"
  )
```

Maybe a frequency chart would be nice,I'll come back to do that later ***(Maybe)***
