---
title: "Holland Portfolio 8: EFA and CFA in R"
---

In my stats courses we've talked a lot about factor analysis — both exploratory (EFA) and confirmatory (CFA). But most of the time we just look at tables of factor loadings and fit indices, which is... not very exciting. This portfolio is a quick exercise in doing both EFA and CFA in R, with an emphasis on making the results actually visible.

## Portfolio Goals

+ Simulate data with a known factor structure
+ Run EFA to "discover" the structure from the data
+ Run CFA to confirm it
+ Visualize the results: loading heatmaps, path diagrams, and fit comparison



## Part 1: Setup and Data Simulation

I'll simulate data with 3 latent factors, each measured by 4 items. The nice thing about simulated data is we know the ground truth, so we can check whether EFA and CFA recover it correctly.

```{r setup, warning=FALSE, message=FALSE}
library(tidyverse)
library(lavaan)
library(psych)
library(semPlot)
library(corrplot)
```

```{r simulate}
set.seed(2026)
n <- 400

# 3 latent factors
F1 <- rnorm(n)
F2 <- rnorm(n)
F3 <- rnorm(n)

# Allow some correlation between factors
F2 <- 0.3 * F1 + sqrt(1 - 0.3^2) * F2
F3 <- 0.2 * F1 + sqrt(1 - 0.2^2) * F3

# 4 items per factor, loadings between 0.5 and 0.8
dat <- tibble(
  x1 = 0.75*F1 + rnorm(n, sd = 0.5),
  x2 = 0.70*F1 + rnorm(n, sd = 0.5),
  x3 = 0.65*F1 + rnorm(n, sd = 0.5),
  x4 = 0.60*F1 + rnorm(n, sd = 0.5),
  x5 = 0.80*F2 + rnorm(n, sd = 0.5),
  x6 = 0.70*F2 + rnorm(n, sd = 0.5),
  x7 = 0.65*F2 + rnorm(n, sd = 0.5),
  x8 = 0.55*F2 + rnorm(n, sd = 0.5),
  x9 = 0.75*F3 + rnorm(n, sd = 0.5),
  x10 = 0.70*F3 + rnorm(n, sd = 0.5),
  x11 = 0.60*F3 + rnorm(n, sd = 0.5),
  x12 = 0.50*F3 + rnorm(n, sd = 0.5)
)

cat("Simulated:", n, "observations,", ncol(dat), "items\n")
```


### Correlation matrix

Before doing any factor analysis, let's just look at the correlations. If the factor structure is real, we should see blocks of higher correlations among items that share a factor.

```{r corr-matrix, fig.width=7, fig.height=6}
cor_mat <- cor(dat)

corrplot(cor_mat, method = "color", type = "lower",
         tl.col = "black", tl.cex = 0.9,
         addCoef.col = "black", number.cex = 0.6,
         col = colorRampPalette(c("#2980B9", "white", "#E74C3C"))(200),
         title = "Correlation Matrix: 12 Items",
         mar = c(0, 0, 2, 0))
```

You can already see the three blocks: x1-x4, x5-x8, x9-x12. That's a good sign — EFA should be able to pick this up.



## Part 2: EFA — Exploratory Factor Analysis

EFA doesn't assume any structure. It looks at the correlations and tries to find the best set of latent factors to explain them. The key decisions are: how many factors, and what rotation.

### 2a: How many factors?

```{r parallel-analysis, fig.width=7, fig.height=5}
# Parallel analysis: compare eigenvalues from data vs. random data
fa.parallel(dat, fa = "fa", main = "Parallel Analysis: How Many Factors?")
```

Parallel analysis compares our data's eigenvalues against what you'd expect from random noise. Factors are retained only if their eigenvalue exceeds the random baseline. It should suggest 3 factors here.


### 2b: Run EFA

```{r efa}
efa_result <- fa(dat, nfactors = 3, rotate = "oblimin", fm = "ml")
print(efa_result, cut = 0.3, sort = TRUE)
```

Oblimin rotation allows factors to be correlated, which is realistic for most psychological constructs. Let's visualize the loadings.

```{r efa-heatmap, fig.width=8, fig.height=5}
# Extract and reshape loadings for ggplot
loadings_df <- as.data.frame(unclass(efa_result$loadings)) %>%
  rownames_to_column("item") %>%
  pivot_longer(-item, names_to = "factor", values_to = "loading")

ggplot(loadings_df, aes(x = factor, y = fct_rev(item), fill = loading)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(loading, 2)), size = 3) +
  scale_fill_gradient2(low = "#2980B9", mid = "white", high = "#E74C3C",
                       midpoint = 0, limits = c(-1, 1)) +
  labs(title = "EFA Factor Loadings (Oblimin Rotation)",
       x = "Factor", y = "Item", fill = "Loading") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

```{r efa-diagram, fig.width=8, fig.height=6}
fa.diagram(efa_result, main = "EFA Factor Structure")
```

If the simulation worked as intended, each item should load primarily on its "true" factor, with small cross-loadings on the other two. The heatmap makes this easy to see at a glance.



## Part 3: CFA — Confirmatory Factor Analysis

Now we pretend we didn't simulate the data. Based on EFA (or theory), we hypothesize a 3-factor structure and test it with CFA.

```{r cfa}
cfa_model <- '
  F1 =~ x1 + x2 + x3 + x4
  F2 =~ x5 + x6 + x7 + x8
  F3 =~ x9 + x10 + x11 + x12
'

cfa_fit <- cfa(cfa_model, data = dat)
summary(cfa_fit, fit.measures = TRUE, standardized = TRUE)
```

```{r cfa-fit}
# Extract fit indices
fit_vals <- fitMeasures(cfa_fit, c("chisq", "df", "pvalue", "cfi", "tli", 
                                     "rmsea", "rmsea.ci.lower", "rmsea.ci.upper", "srmr"))
fit_df <- tibble(
  Index = names(fit_vals),
  Value = round(as.numeric(fit_vals), 3)
)
fit_df
```

<span style="color: deepskyblue;">*(The conventional cutoffs: CFI > .95, TLI > .95, RMSEA < .06, SRMR < .08. With simulated data that follows the model exactly, fit should be excellent. In real data, it's never this clean.)*</span>


### 3a: Path diagram

```{r cfa-sempath, fig.width=9, fig.height=7}
semPaths(cfa_fit,
         whatLabels = "std",
         style = "lisrel",
         layout = "tree2",
         edge.label.cex = 0.8,
         sizeMan = 6, sizeLat = 10,
         residuals = FALSE,
         intercepts = FALSE,
         nCharNodes = 0,
         fade = FALSE,
         edge.color = "gray30",
         mar = c(2, 2, 2, 2))
title("CFA: 3-Factor Model (Standardized Loadings)", line = -1)
```


### 3b: Standardized loadings comparison

```{r loading-comparison, fig.width=9, fig.height=5}
# Extract CFA standardized loadings
std_est <- standardizedSolution(cfa_fit)
cfa_loadings <- std_est %>%
  filter(op == "=~") %>%
  select(item = rhs, factor = lhs, cfa_loading = est.std)

# Compare with EFA primary loadings
efa_primary <- loadings_df %>%
  group_by(item) %>%
  slice_max(abs(loading), n = 1) %>%
  ungroup() %>%
  select(item, efa_loading = loading)

comparison <- cfa_loadings %>%
  left_join(efa_primary, by = "item")

ggplot(comparison, aes(x = efa_loading, y = cfa_loading)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
  geom_point(aes(color = factor), size = 3, alpha = 0.8) +
  geom_text(aes(label = item), nudge_y = 0.02, size = 2.5) +
  scale_color_manual(values = c("F1" = "#E74C3C", "F2" = "#2980B9", "F3" = "#27AE60")) +
  labs(title = "EFA vs CFA Standardized Loadings",
       subtitle = "Points near the diagonal = EFA and CFA agree",
       x = "EFA Primary Loading", y = "CFA Standardized Loading",
       color = "Factor") +
  coord_equal() +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"))
```

When EFA and CFA agree, the points cluster along the diagonal. Deviations suggest that the exploratory solution and the confirmatory model capture slightly different aspects of the data — which is normal, since EFA allows cross-loadings while CFA forces them to zero.


### 3c: What if we get the structure wrong?

Let's see what happens when we specify a wrong model — say, a 2-factor model that lumps F2 and F3 together.

```{r cfa-wrong}
wrong_model <- '
  F1 =~ x1 + x2 + x3 + x4
  F23 =~ x5 + x6 + x7 + x8 + x9 + x10 + x11 + x12
'

wrong_fit <- cfa(wrong_model, data = dat)

# Compare fit
fit_comparison <- tibble(
  Index = c("CFI", "TLI", "RMSEA", "SRMR"),
  Correct_3F = round(as.numeric(fitMeasures(cfa_fit, c("cfi","tli","rmsea","srmr"))), 3),
  Wrong_2F = round(as.numeric(fitMeasures(wrong_fit, c("cfi","tli","rmsea","srmr"))), 3)
)

fit_comparison
```

```{r fit-viz, fig.width=8, fig.height=4}
fit_comparison %>%
  pivot_longer(-Index, names_to = "model", values_to = "value") %>%
  ggplot(aes(x = Index, y = value, fill = model)) +
  geom_col(position = "dodge", width = 0.6, alpha = 0.85) +
  geom_text(aes(label = value), position = position_dodge(width = 0.6),
            vjust = -0.3, size = 3) +
  scale_fill_manual(values = c("Correct_3F" = "#27AE60", "Wrong_2F" = "#E74C3C"),
                    labels = c("3-Factor (correct)", "2-Factor (wrong)")) +
  labs(title = "Model Fit Comparison: Correct vs. Misspecified Model",
       y = "Fit Index Value", x = NULL, fill = "") +
  theme_minimal() +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

The correct 3-factor model should show better fit on every index. This is the whole point of CFA: testing whether your theoretical structure actually fits the data better than alternatives.


<span style="color: deepskyblue;">*Bravo!!! EFA finds the structure, CFA confirms it, and the misspecified model proves why getting the structure right matters. The heatmap and the EFA-vs-CFA scatter plot are my favorite parts — they make factor loadings way more digestible than a wall of numbers.*</span>
